---
title: "Reproducible Research Week 4 Course Project 2"
output: html_document
---

```{r setwd}
library(ggplot2)  
library(plyr) 
library(dplyr)
setwd("C:/Users/CathyHoi/Dropbox/PhD docs/Certificates/Reproducible Research Course Project 2")
```

Read and process data
```{r process}
data<-read.csv("repdata_data_StormData.csv")

## Processing data
dim(data)

str(data)

## Extracting variables of interest for analysis
selected <- c( "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
selecteddata <- data[, selected]

## Check the first and last few rows 
head(selecteddata)
tail(selecteddata)

## Check the missing values in the dataset
sum(is.na(selecteddata$FATALITIES))
sum(is.na(selecteddata$INJURIES))
sum(is.na(selecteddata$PROPDMG))
sum(is.na(selecteddata$CROPDMG))
sum(is.na(selecteddata$PROPDMGEXP))
sum(is.na(selecteddata$CROPDMGEXP))
```

## Group any event that contains "WIND" into one type, and the same for all variables  

```{r group}
## List the first 10 most frequent events 
sort(table(selecteddata$EVTYPE), decreasing = TRUE) [1:10]

selecteddata$EVENT <- "OTHER"

selecteddata$EVENT[grep("HAIL", selecteddata$EVTYPE, ignore.case = TRUE)] <- "HAIL"
selecteddata$EVENT[grep("HEAT", selecteddata$EVTYPE, ignore.case = TRUE)] <- "HEAT"
selecteddata$EVENT[grep("FLOOD", selecteddata$EVTYPE, ignore.case = TRUE)] <- "FLOOD"
selecteddata$EVENT[grep("WIND", selecteddata$EVTYPE, ignore.case = TRUE)] <- "WIND"
selecteddata$EVENT[grep("STORM", selecteddata$EVTYPE, ignore.case = TRUE)] <- "STORM"
selecteddata$EVENT[grep("SNOW", selecteddata$EVTYPE, ignore.case = TRUE)] <- "SNOW"
selecteddata$EVENT[grep("TORNADO", selecteddata$EVTYPE, ignore.case = TRUE)] <- "TORNADO"
selecteddata$EVENT[grep("WINTER", selecteddata$EVTYPE, ignore.case = TRUE)] <- "WINTER"
selecteddata$EVENT[grep("RAIN", selecteddata$EVTYPE, ignore.case = TRUE)] <- "RAIN"
```

```{r list}
## List them again
sort(table(selecteddata$EVENT), decreasing = TRUE)
sort(table(selecteddata$PROPDMGEXP), decreasing = TRUE)[1:10]
sort(table(selecteddata$CROPDMGEXP), decreasing = TRUE)[1:10]
```

## Create new variables
```{r new variables}
selecteddata$PROPDMGEXP <- as.character(selecteddata$PROPDMGEXP)
selecteddata$PROPDMGEXP[is.na(selecteddata$PROPDMGEXP)] <- 0 
selecteddata$PROPDMGEXP[!grepl("K|M|B", selecteddata$PROPDMGEXP, ignore.case = TRUE)] <- 0 
selecteddata$PROPDMGEXP[grep("K", selecteddata$PROPDMGEXP, ignore.case = TRUE)] <- "3"
selecteddata$PROPDMGEXP[grep("M", selecteddata$PROPDMGEXP, ignore.case = TRUE)] <- "6"
selecteddata$PROPDMGEXP[grep("B", selecteddata$PROPDMGEXP, ignore.case = TRUE)] <- "9"
selecteddata$PROPDMGEXP <- as.numeric(as.character(selecteddata$PROPDMGEXP))
selecteddata$property.damage <- selecteddata$PROPDMG * 10^selecteddata$PROPDMGEXP
selecteddata$CROPDMGEXP <- as.character(selecteddata$CROPDMGEXP)
selecteddata$CROPDMGEXP[is.na(selecteddata$CROPDMGEXP)] <- 0 
selecteddata$CROPDMGEXP[!grepl("K|M|B", selecteddata$CROPDMGEXP, ignore.case = TRUE)] <- 0 
selecteddata$CROPDMGEXP[grep("K", selecteddata$CROPDMGEXP, ignore.case = TRUE)] <- "3"
selecteddata$CROPDMGEXP[grep("M", selecteddata$CROPDMGEXP, ignore.case = TRUE)] <- "6"
selecteddata$CROPDMGEXP[grep("B", selecteddata$CROPDMGEXP, ignore.case = TRUE)] <- "9"
selecteddata$CROPDMGEXP <- as.numeric(as.character(selecteddata$CROPDMGEXP))
selecteddata$crop.damage <- selecteddata$CROPDMG * 10^selecteddata$CROPDMGEXP
```

## List the first 10 most appearred values of property and crop damage 
```{r list 2}
sort(table(selecteddata$property.damage), decreasing = TRUE)[1:10]

sort(table(selecteddata$crop.damage), decreasing = TRUE)[1:10]
```

Answer for the question: Across the United States, which types of events (as indicated in the EVTYPE\color{red}{\verb|EVTYPE|}EVTYPE variable) are most harmful with respect to population health?

```{r quesion1}
# Both fatalities and injuries 
FatalInj <- ddply(selecteddata, .(EVENT), summarize, Total = sum(FATALITIES + INJURIES,  na.rm = TRUE))
FatalInj$type <- "Fatalities and Injuries"

# Fatalities only
Fatal <- ddply(selecteddata, .(EVENT), summarize, Total = sum(FATALITIES, na.rm = TRUE))
Fatal$type <- "fatalities"

# Injuries only
Inj <- ddply(selecteddata, .(EVENT), summarize, Total = sum(INJURIES, na.rm = TRUE))
Inj$type <- "injuries"

# Combine all into one variable
Health <- rbind(Fatal, Inj)

HarmHealth <- join (Fatal, Inj, by="EVENT", type="inner")
HarmHealth
```

Plot 1 for the question
```{r plot1}
# Transform "EVENT" to factor variable
Health$EVENT <- as.factor(Health$EVENT)

# Plot fatalities and injuries 
Plot1 <- ggplot(Health, aes(x = EVENT, y = Total, fill = type)) + geom_bar(stat = "identity") +
                coord_flip() +
                xlab("Event Type") + 
                ylab("Total number of health impact") +
                ggtitle("Weather event types impact on public health") +
                theme(plot.title = element_text(hjust = 0.5))

Plot1  
```

Answer for the question: Across the United States, which types of events have the greatest economic consequences?

```{r question2}
# Both property damage and crop damage 
PropCrop <- ddply(selecteddata, .(EVENT), summarize, Total = sum(property.damage + crop.damage,  na.rm = TRUE))
PropCrop$type <- "property and crop damage"

# Property damage only
Prop <- ddply(selecteddata, .(EVENT), summarize, Total = sum(property.damage, na.rm = TRUE))
Prop$type <- "property"

# Crop damage only
Crop <- ddply(selecteddata, .(EVENT), summarize, Total = sum(crop.damage, na.rm = TRUE))
Crop$type <- "crop"

# Combine all into one variable
Economic <- rbind(Prop, Crop)


HarmEconomic <- join (Prop, Crop, by="EVENT", type="inner")
HarmEconomic
```

Plot 2 for the question

```{r plot2}
# Transform "EVENT" to factor variable
Economic$EVENT <- as.factor(Economic$EVENT)

# Plot property and crop damage 
Plot2 <- ggplot(Economic, aes(x = EVENT, y = Total, fill = type)) + geom_bar(stat = "identity") +
                coord_flip() +
                xlab("Event Type") + 
                ylab("Total damage in dollars") +
                ggtitle("Weather event types impact on property and crop damage") +
                theme(plot.title = element_text(hjust = 0.5))

Plot2
```
